---
title: "Base R Plots"
author: "Joey Miranda"
date: "9/16/2014"
output:
  word_document: null
  pdf_document: default
  html_document: default
email: jmiranda@zagmail.gonzaga.edu
abstract: The objective of this document's template is to demonstrate some basics
  of the combination of R and markdown and how they can be knitted together using
  the knitr package (via the RStudio IDE) to produce beautiful docs/reports.
---

#Purpose

Assignment 3: This document shows various strings of code with the coresponding graphs to get a better understanding of how to use the graphs with RMarkdown

#Data
```{r}

## Simulate some data

## 3 Factor Variables
FacVar1=as.factor(rep(c("level1","level2"),25))
FacVar2=as.factor(rep(c("levelA","levelB","levelC"),17)[-51])
FacVar3=as.factor(rep(c("levelI","levelII","levelIII","levelIV"),13)[-c(51:52)])

## 4 Numeric Variables
set.seed(123)
NumVar1=round(rnorm(n=50,mean=1000,sd=50),digits=2) ## Normal distribution
set.seed(123)
NumVar2=round(runif(n=50,min=500,max=1500),digits=2) ## Uniform distribution
set.seed(123)
NumVar3=round(rexp(n=50,rate=.001)) ## Exponential distribution
NumVar4=2001:2050

simData=data.frame(FacVar1,FacVar2,FacVar3,NumVar1,NumVar2,NumVar3,NumVar4)

```

# Plotting one Variable: Numeric Variable

```{r}
plot(simData$NumVar1,type="o") ## Index plot, 'type="o"'adds the line connecting each data point, 'plot("dataset")' is one of the most common of RMardown code for graphing a numeric variable dataset
```

```{r}
hist(simData$NumVar1) ## Histogram, replacing 'plot' with 'hist' using the same data set will create a Histogram instead of a data plot graph
```

```{r}
plot(density(simData$NumVar1)) ## Kernel density plot: adding '(denisty("dataset"))' to data plot will create a Kernel density play,used to see a distribution of a variable 
```

```{r}
boxplot(simData$NumVar1) ## box plot: adding 'box' ouside of plot on a plot graph will create a box chart, used to visualize 25-75% distribution and also the remaining distribution extremes
```

# Plotting one variable: Factor variable

```{r}
plot(simData$FacVar3) ## bar plot: stickig to the plot style but using a data set with factor variables will automatically create the most logical graph with is a bar plot
```

```{r}
## pie chart - Not the best graph --- use with caution
counts=table(simData$FacVar3) ## get counts: this seaches the dataset for how many variables it is using
labs=paste(simData$FacVar3,counts)## create labels: this will take the names of the variables and attach them to each variable from 'counts'
pie(counts,labels=labs) ## plot: this will plot the pie chart using the variables from 'counts' and display the lables from 'labs'
```

#Two Variables: Two Numeric Variables

```{r}
plot(simData$NumVar1,type="o",ylim=c(0,max(simData$NumVar1,simData$NumVar2)))## index plot with one variable: creates a normal lined plot chart but extendes the y limit to fit another data set as well
lines(simData$NumVar2,type="o",lty=2,col="red")## add another variable: addes the second dataset over the first one and colors it red
```

```{r}
## Density plots
dv1=density(simData$NumVar1) 
dv2=density(simData$NumVar2) ##these assign each density chart to a variable 'dv1, dv2'
plot(range(dv1$x, dv2$x),range(dv1$y, dv2$y), type = "n", xlab = "NumVar1(red) and NumVar2 (blue)",ylab = "Density") ##this plots each density chart on one graph and identifies each line with a color in the x axis and marks density on the y axis
lines(dv1, col = "red")
lines(dv2, col = "blue") ## these change each density plot to a designated color
```

```{r}
plot(simData$NumVar1,simData$NumVar2) ## scatterplots: this is a simple scatterplot, just use the plot function and use two numeric variable sets seperated with a comma
```

#Two Variables: Two Factor Variables

```{r}
plot(table(simData$FacVar2,simData$FacVar3)) ## Mosaic plot: using the plot function and adding 'table("datasets")' to it while using two factor variable sets seperated by a comma, this compares relationships amoung all the variables
```

```{r}
## barplots
bartable=table(simData$FacVar2,simData$FacVar3) ## get the cross tab: this grabs the data sets identifies all the points, 'bartable' will be used for the next as a variable for the next two graphs as well
barplot(bartable,beside=TRUE, legend=levels(unique(simData$FacVar2))) ## plot: this plots 'bartable' placing each variable in a set next to each other('beside=TRUE') and creates a legend using the dataset to identify each bar and relate it
```

```{r}
barplot(bartable, legend=levels(unique(simData$FacVar2))) ## stacked: simial to the last barplot but it stacks the variables, note no 'beside=TRUE' is the only difference
```

```{r}
barplot(prop.table(bartable,2)*100, legend=levels(unique(simData$FacVar2))) ## stacked 100%: also similar to the last one but sets the variables out of 100% ['prop.table(variable,index margin)']
```

#Two Variables: One Factor and One Numeric

```{r}
## Box plots for the numeric variable over the levels of the factor variable
plot(simData$FacVar1,simData$NumVar1)  ## simply using a data set of factor variables and numeric variables will create a box plot comparing each factor variable
```

```{r}
## density plot of numeric variables across multiple levels of the factor variables
level1=simData[simData$FacVar1=="level1",] 
level2=simData[simData$FacVar1=="level2",] ##assigns and names each factor variable dataset to a certain name 'level1, level2'

dv3=density(level1$NumVar1)
dv4=density(level2$NumVar1) ##assigns a density plot from a numeric variable set 'NumVar1' across both variable sets 'level 1,2'

plot(range(dv3$x, dv4$x),range(dv3$y, dv4$y), type = "n", xlab = "NumVar1 at Level1 (red) and NumVar1 at Level2 (blue)",ylab = "Density")
lines(dv3, col = "red")
lines(dv4, col = "blue") ##plots the graph, and marks each axis and colors each data line
```

```{r}
## Mean of one numeric variable over levels of one factor variable
meanagg=aggregate(simData$NumVar1, list(simData$FacVar3), mean) ## using 'aggregate(dataset)' will split the data into subsets, lists 'list()' them according to each factor variable, and finds the mean 'mean' of each numeric veriable in relation to the factor variable

dotchart(meanagg$x,labels=meanagg$Group.1) ## Dot Chart: plots the chart along with labeling the y axis with each factor variable
```

```{r}
barplot(meanagg$x,names.arg=meanagg$Group.1)## Bar plot: using the same variables 'meanagg$x, meanagg$Group.1' but instead of dots, it creates a bar chart. note: this is not an appropriate method for displaying a mean since the bars are too difficult to discern the differences
```

#Three Variables: Three Factor Variables

```{r}
par(mfrow=c(1,2)) ##this makes it so both graphs are placed next to each other to help compare easier

bar1table=table(level1$FacVar2,level1$FacVar3)
barplot(bar1table,beside=TRUE, main="FacVar1=level1")

bar2table=table(level2$FacVar2,level2$FacVar3)
barplot(bar2table,beside=TRUE, main="FacVar1=level2", legend=levels(unique(level2$FacVar2))) ##these create two barplots that display the levels or two different factor variables 'FacVar2,3' to another differnet factor variable 'level1,2'
```

#Three Variables: One Numeric and Two Factor Variables

```{r}
par(mfrow=c(1,1))
## boxplot of NumVar1 over an interaction of 6 levels of the combination of FacVar1 and FacVar2
boxplot(NumVar1~interaction(FacVar1,FacVar2),data=simData) ##creates a box plot of six boxes, each showing the relation of the numeric variable 'NumVar1' to both factor variables 'FacVar1,2','interaction'is used to find the interaction between the numeric variable and each factor variable
```

```{r}
## Mean of 1 Numeric over levels of two factor vars
meanaggg=aggregate(simData$NumVar1, list(simData$FacVar1,simData$FacVar2), mean)
meanaggg=meanaggg[order(meanaggg$Group.1),]
meanaggg$color[meanaggg$Group.2=="levelA"] = "red"
meanaggg$color[meanaggg$Group.2=="levelB"] = "blue"
meanaggg$color[meanaggg$Group.2=="levelC"] = "darkgreen"  

dotchart(meanaggg$x,labels=meanaggg$Group.2, groups=meanaggg$Group.1,color=meanaggg$color) ## dotchart: similar to the last dot chart displaying a mean, this one takes 'NumVar1' and finds its mean in relation to both 'FacVar1,2' and places it on one graph
```

```{r}
interaction.plot(meanaggg$Group.2,meanaggg$Group.1,meanaggg$x,type="b", col=c(1:2),pch=c(18,24)) ## interaction plot - line plots of means: creates a line chart that also relates the mean of a numeric variable to two factor variables, and places the means at data points with a line connecting each, it is color coded and each data point has a different symbol based on which factor variable it is for
```

```{r}
## some a bar plot
par(mfrow=c(1,2))

level1=meanaggg[meanaggg$Group.1=="level1",]
level2=meanaggg[meanaggg$Group.1=="level2",]

barplot(level1$x,names.arg=level1$Group.2, main="FacVar1=level1")
barplot(level2$x,names.arg=level2$Group.2, main="FacVar1=level2") ##very smilar to the last few graphs and the numeric mean bar chart, but uses one more factor variable, again note this is not ideal for displaying means
```

#Three Variables: Two Numeric and One Factor Variables

```{r}
## Scatter plot with color identifying the factor variable
par(mfrow=c(1,1))
plot(simData$NumVar1,simData$NumVar2, col=simData$FacVar1)
legend("topright",levels(simData$FacVar1),fill=simData$FacVar1) ##this plots points from both numeric variables in relation to the factor variable and each factor variable is color coded and uses the legend identify each factor
```

#Three Variables: Three Numeric Variables

```{r}
## NumVar4 is 2001 through 2050... possibly, a time variable - use that as the x-axis: this places two different numeric variables 'NumVar1,2' on a grid, each one is connected with a line and color coded, both variables are in relation to the third numeric variable 'NumVar4' which is places on the x axis
plot(simData$NumVar4,simData$NumVar1,type="o",ylim=c(0,max(simData$NumVar1,simData$NumVar2)))## join dots with lines

lines(simData$NumVar4,simData$NumVar2,type="o",lty=2,col="red")## add another line
```

```{r}
## Bubble plot - scatter plot of NumVar1 and NumVar2 with individual observations sized by NumVar3: this places numeric data on a grid relating two different numeric variables 'NumVar1,2', while identifying with a third variable 'NumVar3'
radius <- sqrt( simData$NumVar3/ pi ) ##sets the radius of each variable from NumVar3
symbols(simData$NumVar1,simData$NumVar2,circles=radius, inches=.25,fg="white", bg="red", main="Sized by NumVar3")
```

#Scatterplot Matrix of all Numeric Vars, colored by a Factor variable

```{r}
pairs(simData[,4:7], col=simData$FacVar1) ##creates a scatter plot of every numeric variable compared to each numeric variable in relation to one factor variable 'FacVar1'
```